\pagelayout{margin}
\chapter*{Acknowledgments}


I hope the reader appreciates the hard work that has ultimately produced this thesis, which is the result of the direct and indirect effort of many people. 

First and foremost, I am infinitely grateful for the opportunities, guidance, help and trust my supervisors have offered me. I have been extremely lucky to have such passionate, knowledgeable and caring people directing this work. These four years have allowed me to be creative, to learn much more and to go beyond even what I expected. Thank you, dad. Thank you, Paco.

Learning is one of my passions. I am thankful for all the good teachers I have had, they are the motivation that students need to strive for the best. I want to take a moment to thank and remember two of them who sadly are no longer with us. With Manuel Arroquia I learned not to accept something as truth without finding a reasoning. Julio Ortega showed me that there is joy in teaching and caring about your students. The curiosity you inspired in your students and the aspiration for excellence you inspired in your colleagues live on.

Big thanks to my friends in the SCI2S group and at the CITIC for making these years fun, making me see things from other perspectives and being there to support each other.

It is fairly uncommon that one of your supervisors also happens to be your father. Because of this, the rest of my family had to bear with many boring conversations at home when work topics inevitably came up. Still, they have always been there to share a good time with us. Mom, Alex, thank you for your endless patience and support.

Finally, I have to express my gratitude to the person who has always been next to me during these four years, who has lifted my spirits more times than I could count and has shown me nothing but love. Thank you from the bottom of my heart, Esperanza.

This work is a reality thanks to all of you.

\pagelayout{wide}